\LaTeXe\ guide for the Journal of Functional Programming

### 2019-03-05 papers/jfp/1-intro.tex
Add first JFP version

Build systems (such as \Make) are big, complicated, and used by every
software developer on the planet.  But they are a sadly unloved part
of the software ecosystem, very much a means to an end, and seldom the
focus of attention.
% Rarely do people ask questions like ``What does it mean for my build
% system to be correct?'' or ``What are the trade-offs between different
% approaches?''.  Moreover,
For years \Make dominated, but more recently the challenges of scale have driven
large software firms like Microsoft, Facebook, and Google to develop their own
build systems, exploring new points in the design space. These complex build
systems use subtle algorithms, but they are often hidden away, and not the
object of study.

In this paper we offer a general framework in which to understand and compare
build systems, in a way that is both abstract (omitting incidental detail)
and yet precise (implemented as Haskell code). Specifically we make these
contributions:
\begin{itemize}
\item Build systems vary on many axes, including: static vs dynamic
  dependencies; local vs cloud; deterministic vs non-deterministic build rules;
  support for early cutoff; self-tracking build systems; and the type of
  persistent build information. In~\S\ref{sec-background} we identify some key
  properties, illustrated by four carefully-chosen build systems.

@@

\item We identify two key design choices that are typically deeply wired into
  any build system: \emph{the order in which dependencies are
  built}~(\S\ref{sec-dependency-orderings}) and \emph{whether or not a
  dependency is (re-)built}~(\S\ref{sec-out-of-date}). These choices turn out to
  be orthogonal, which leads us to a new classification of the design
  space~(\S\ref{sec-design-space}).

>>>>

Build systems (such as \Make) are big, complicated, and used by every
software developer on the planet.  But they are a sadly unloved part
of the software ecosystem, very much a means to an end, and seldom the
focus of attention.
% Rarely do people ask questions like ``What does it mean for my build
% system to be correct?'' or ``What are the trade-offs between different
% approaches?''.
For years \Make dominated, but more recently the challenges of scale have driven
large software firms like Microsoft, Facebook and Google to develop their own
build systems, exploring new points in the design space. These complex build
systems use subtle algorithms, but they are often hidden away, and not the
object of study.

In this paper we offer a general framework in which to understand and compare
build systems, in a way that is both abstract (omitting incidental detail)
and yet precise (implemented as Haskell code). Specifically we make these
contributions:
\begin{itemize}
\item Build systems vary on many axes, including: static vs dynamic
  dependencies; local vs cloud; deterministic vs non-deterministic build tasks;
  support for early cutoff; self-tracking build systems; and the type of
  persistent build information. In~\S\ref{sec-background} we identify some key
  properties, illustrated by four carefully-chosen build systems.

@@

\item We identify two key design choices that are typically deeply wired into
  any build system: \emph{the order in which tasks are
  built}~(\S\ref{sec-dependency-orderings}) and \emph{whether or not a
  task is (re-)built}~(\S\ref{sec-out-of-date}). These choices turn out to
  be orthogonal, which leads us to a new classification of the design
  space~(\S\ref{sec-design-space}).

### 2019-03-05 papers/jfp/2-background.tex
Add first JFP version

\definition[Minimality]{A build system is \emph{minimal} if it executes tasks at
most once per build and only if they transitively depend on inputs that changed
since the previous build.}\label{def-minimal}
\vspace{2mm}

@@

\begin{table}[h]
\vspace{-2mm}
\caption{Build system differences.\label{tab-summary}}
\vspace{-4mm}
\smaller
\centering
\begin{tabular}{l||l|l||l|c|c|c}
\hline
$\!$Build system$\!$& Persistent build information & Scheduler   & Dependencies    & Minimal & Cutoff & Cloud$\!$\\\hline
$\!$\Make       $\!$& File modification times      & Topological & Static          & Yes     & No     & No   $\!$\\
$\!$\Excel      $\!$& Dirty cells, calc chain      & Restarting  & Dynamic         & No      & No     & No   $\!$\\
$\!$\Shake      $\!$& Previous dependency graph    & Suspending  & Dynamic         & Yes     & Yes    & No   $\!$\\
$\!$\Bazel      $\!$& Cloud cache, command history & Restarting  & Dynamic$^{(*)}$ & No      & Yes    & Yes  $\!$\\\hline
\end{tabular}
\vspace{2mm}
$^{(*)}$At present, user-defined build rules cannot have dynamic dependencies.
\vspace{-8mm}
\end{table}

>>>>

\textbf{Definition} (Minimality). A build system is \emph{minimal} if it
executes tasks at most once per build and only if they transitively depend on
inputs that changed since the previous build.\label{def-minimal}
\vspace{2mm}

@@

\begin{table}[h]
\vspace{-2mm}
\caption{Build system differences.\label{tab-summary}}
\vspace{-4mm}
% \smaller
\centering
\begin{tabular}{l||l|l||l|c|c|c}
\hline
$\!$Build system$\!$& Persistent build information & Scheduler   & Dependencies    & Minimal & Cutoff & Cloud$\!$\\\hline
$\!$\Make       $\!$& File modification times      & Topological & Static          & Yes     & No     & No   $\!$\\
$\!$\Excel      $\!$& Dirty cells, calc chain      & Restarting  & Dynamic         & No      & No     & No   $\!$\\
$\!$\Shake      $\!$& Previous dependency graph    & Suspending  & Dynamic         & Yes     & Yes    & No   $\!$\\
$\!$\Bazel      $\!$& Cloud cache, command history & Restarting  & Dynamic$^{(*)}$ & No      & Yes    & Yes  $\!$\\\hline
\end{tabular}
\vspace{2mm}
$^{(*)}$At present, user-defined build rules cannot have dynamic dependencies.
\vspace{-8mm}
\end{table}

### 2019-03-05 papers/jfp/3-abstractions.tex
Add first JFP version

\vspace{-1mm}
\definition[Correctness]{Suppose \hs{build} is a build system, \hs{tasks} is a
build task description, \hs{key} is a target key, \hs{store} is an initial
store, and \hs{result} is the store produced by running the build system with
parameters \hs{tasks}, \hs{key} and \hs{store}. Or, using the precise language of
our abstractions:

@@

    \item The \hs{result} is \emph{consistent} with the \hs{tasks}, i.e.
    for all non-input keys \hs{k}~$\in$~$O$, the result of recomputing the
    corresponding \hs{task} matches the value stored in the \hs{result}:
    \[
    \hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}.
    \]
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{tasks}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{2mm}

@@

\noindent
Notice that these calls to \hs{dependencies} do no actual computation (in this
case, spreadsheet arithmetic). They cannot: we are not supplying a store or any
input numbers. So, through the wonders of polymorphism, we are able to extract
the dependencies of the spreadsheet formula, and to do so efficiently, simply by
running its code in a different \hs{Applicative}! This is not new, for example
see~\citet{free-applicatives}, but it is extremely cool. We will see a practical
use for \hs{dependencies} when implementing applicative build systems,
see~\S\ref{sec-implementation-make}.

>>>>

% \vspace{-1mm}
\textbf{Definition}(Correctness). Suppose \hs{build} is a build system,
\hs{tasks} is a build task description, \hs{key} is a target key, \hs{store} is
an initial store, and \hs{result} is the store produced by running the build
system with parameters \hs{tasks}, \hs{key} and \hs{store}. Or, using the
precise language of our abstractions:

@@

    \item The \hs{result} is \emph{consistent} with the \hs{tasks}, i.e.
    for all non-input keys \hs{k}~$\in$~$O$, the result of recomputing the
    corresponding \hs{task} matches the value stored in the \hs{result}:
    \[
    \hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}.
    \]
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{tasks}, \hs{key} and \hs{store}.

\label{def-correct}
\vspace{2mm}

@@

\noindent
Notice that these calls to \hs{dependencies} do no actual computation (in this
case, spreadsheet arithmetic). They cannot: we are not supplying a store or any
input numbers. So, through the wonders of polymorphism, we are able to extract
the dependencies of the spreadsheet formula, and to do so efficiently, simply by
running its code in a different \hs{Applicative}! This is not new, for example
see~\cite{free-applicatives}, but it is extremely cool. We will see a practical
use for \hs{dependencies} when implementing applicative build systems,
see~\S\ref{sec-implementation-make}.

### 2019-03-05 papers/jfp/6-engineering.tex
Add first JFP version

\item The \hs{suspending} scheduler can be made parallel by starting static
dependencies of a \hs{Task} in parallel, while dynamic dependencies are being
resolved, using the strategy described by~\citet{marlow2014haxl}.
\end{enumerate}

@@

\noindent
Such tasks can be modelled in our framework by adjusting the correctness
definition (\S\ref{sec-build-correctness}): instead of requiring that the
\hs{result}ing value \emph{equals} the one obtained by recomputing the \hs{task}:
\[
\hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}\text{,}
\]
\noindent
we now require that \hs{result} contains \emph{one possible result of
recomputing} the \hs{task}:
\[
\hs{getValue}~\hs{k}~\hs{result}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result}\text{,}
\]
where
\hs{computeND}~\hs{::}~\hs{Task}~\hs{MonadPlus}~\hs{k}~\hs{v}~\hs{->}~\hs{Store}~\hs{i}~\hs{k}~\hs{v}~\hs{->}~\hs{[@@v]}
returns the list of all possible results of the \hs{task} instead of just one
value (`ND' stands for `non-deterministic').

@@

To legitimise shallow builds, we need to relax the correctness
Definition~\ref{def-correct} as follows. Let the \hs{shallow} store correspond
to the result of a shallow build. Then \hs{shallow} is correct, if \emph{there
exists} a \hs{result} which satisfies all requirements of the
Definition~\ref{def-correct}, \emph{such that} \hs{shallow} agrees with the
\hs{result} on all the input keys \hs{k}~$\in$~$I$:
\[
\hs{getValue}~\hs{k}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result}\text{,}
\]
and on the target \hs{key}:
\[
\hs{getValue}~\hs{key}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{key}~\hs{result}\text{.}
\]

@@

For examples like \LaTeX~we consider the proper encoding to not be circular
tasks, but a series of iterative steps, as described
by~\citet{shake-fixed-point}. It is important that the number of executions is
bounded, otherwise the build system may not terminate (a legitimate concern
with \LaTeX, which can be put into a situation where it is bistable or diverging
over multiple executions). The examples in \Excel tend to encode either mutable
state, or recurrence relations. The former is only required because \Excel
inherently lacks the ability to write mutable state, and the latter is probably
better solved using explicit recurrence formulae.

>>>>

\item The \hs{suspending} scheduler can be made parallel by starting static
dependencies of a \hs{Task} in parallel, while dynamic dependencies are being
resolved, using the strategy described by~\cite{marlow2014haxl}.
\end{enumerate}

@@

\noindent
Such tasks can be modelled in our framework by adjusting the correctness
definition (\S\ref{sec-build-correctness}): instead of requiring that the
\hs{result}ing value \emph{equals} the one obtained by recomputing the \hs{task}:
\[
\hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result} %\text{,}
\]
\noindent
we now require that \hs{result} contains \emph{one possible result of
recomputing} the \hs{task}:
\[
\hs{getValue}~\hs{k}~\hs{result}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result} %\text{,}
\]
where
\hs{computeND}~\hs{::}~\hs{Task}~\hs{MonadPlus}~\hs{k}~\hs{v}~\hs{->}~\hs{Store}~\hs{i}~\hs{k}~\hs{v}~\hs{->}~\hs{[@@v]}
returns the list of all possible results of the \hs{task} instead of just one
value (`ND' stands for `non-deterministic').

@@

To legitimise shallow builds, we need to relax the correctness
Definition~\ref{def-correct} as follows. Let the \hs{shallow} store correspond
to the result of a shallow build. Then \hs{shallow} is correct, if \emph{there
exists} a \hs{result} which satisfies all requirements of the
Definition~\ref{def-correct}, \emph{such that} \hs{shallow} agrees with the
\hs{result} on all the input keys \hs{k}~$\in$~$I$:
\[
\hs{getValue}~\hs{k}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result} % \text{,}
\]
and on the target \hs{key}:
\[
\hs{getValue}~\hs{key}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{key}~\hs{result} % \text{.}
\]

@@

For examples like \LaTeX~we consider the proper encoding to not be circular
tasks, but a series of iterative steps, as described
by~\cite{shake-fixed-point}. It is important that the number of executions is
bounded, otherwise the build system may not terminate (a legitimate concern
with \LaTeX, which can be put into a situation where it is bistable or diverging
over multiple executions). The examples in \Excel tend to encode either mutable
state, or recurrence relations. The former is only required because \Excel
inherently lacks the ability to write mutable state, and the latter is probably
better solved using explicit recurrence formulae.

### 2019-03-05 papers/jfp/8-conclusions.tex
Add first JFP version

We have investigated multiple build systems, showing how their properties are consequences of two implementation choices: what order you build in and whether you decide to rebuild. By first decomposing the pieces, we show how to recompose the pieces to find new points in the design space. In particular, a simple recombination leads to a design for a monadic suspending cloud build system. Armed with that blueprint we hope to actually implement such a system as future work.

>>>>

We have investigated multiple build systems, showing how their properties are consequences of two implementation choices: what order you build in and how you decide whether to rebuild. By first decomposing the pieces, we show how to recompose the pieces to find new points in the design space. In particular, a simple recombination leads to a design for a monadic suspending cloud build system. Armed with that blueprint we hope to actually implement such a system as future work.

### 2019-03-05 papers/jfp/9-appendix.tex
Add first JFP version

\begin{minted}[xleftmargin=10pt]{haskell}
inputs :: Eq k => Task Monad k v -> Store i k v -> k -> [k]
inputs task store key = filter (isInput task) (closure deps key)
  where
    deps k = maybe [] snd (track task store k)

>>>>

\begin{minted}[xleftmargin=10pt]{haskell}
inputs :: Eq k => Task Monad k v -> Store i k v -> k -> [k]
inputs task store key = filter (isInput task) (closure deps key)
  where
    deps k = maybe [] snd (track task (\k -> getValue k store) k)
